---
title: "HMP_FinalProject_III"
author: "Priya, Ayo"
date: "2023-04-04"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Setup

## Load libraries

```{r setup}
library(readr)
library(tidyverse)
library(ggplot2)
library(readxl)
library(flextable)
library(DataExplorer)
```

## Load data

```{r echo=TRUE}
#Loading datasets
OMEGA <- read.csv("~/Documents/UMICH/courses/UM_Winter_2023/HMP 669/Group Project/OMEGA_filtered.csv")
crit = read_csv('CRIT_cleaned.csv')
# crit_other_services = read_xlsx('CRIT2_other_services.xlsx')
#omega_px_filtered = read_csv('OME_PX_filtered.csv')
MDC_List <- read.csv("~/Documents/UMICH/courses/UM_Winter_2023/HMP 669/Group Project/MDC_List.csv")
omega_other_services = read_csv('OMEGA_other_services.csv')




```




# Data cleaning


## Unique patients

```{r}
OMEGA <- OMEGA %>% 
  distinct(PAT_NO, .keep_all = TRUE) %>% 
  mutate(MAJ_DX_CAT = recode(MAJ_DX_CAT, 
                             `5` = "05 - Cardiology",
                             `8` = "08 - Orthopedic"))
```



# Questions

## 1

Top disease entities treated by each group (Cardiac and Orthopedic) as well as the number of patients in each of these disease entities

### Number of patients in each group

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>% 
  dplyr::summarise(N = n()) %>% 
  flextable()
```


```{r}
OMEGA %>% 
  group_by(MAJ_DX_CAT, Description.1) %>% 
  dplyr::summarise(N = n()) %>% 
  group_by(MAJ_DX_CAT) %>% 
  top_n(5) %>% 
  arrange(-N) %>% 
  flextable()
```


```{r}
OMEGA %>% 
  group_by(MAJ_DX_CAT, Description.1) %>% 
  dplyr::summarise(N = n()) %>% 
  group_by(MAJ_DX_CAT) %>% 
  top_n(5) 
```



## Total charges in each group?

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>%
  summarise(
    `Total Cost` = sum(TOT_COST, na.rm = TRUE) / 1000000,
    `Total Pay` = sum(TOT_PAY, na.rm = TRUE) / 1000000
  ) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  flextable()
```


## Bar chart

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>%
  dplyr::summarise(
    `Total Cost` = sum(TOT_COST, na.rm = TRUE) / 1000000,
    `Total Revenue` = sum(TOT_PAY, na.rm = TRUE) / 1000000
  ) %>% 
  pivot_longer(!MAJ_DX_CAT) %>% 
  ggplot(aes(x = MAJ_DX_CAT, y = value, fill = name)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "$ Million", fill = NULL) +
  theme_bw() +
  theme(text = element_text(size = 16))
  geom_text(aes(label = n), hjust = 1.3, size = 3.5)
#slight loss in ortho
#profitability 

```

## Profitibility

### Histograms

```{r}
OMEGA %>%
  mutate(Profit = TOT_PAY - TOT_COST) %>% 
  ggplot(aes(x = Profit)) +
  geom_histogram(fill = "orange") +
  geom_vline(xintercept = 0, lty = 2, size = 1, color = "red") +
  theme_test() +
  facet_wrap(~MAJ_DX_CAT, scales = "free_x")
  geom_text(aes(label = n), hjust = 1.3, size = 3.5)
```


```{r}
p <- OMEGA %>%
  mutate(Profit = TOT_PAY - TOT_COST) %>% 
  group_by(MAJ_DX_CAT) %>%
  dplyr::summarise(
    `Total Profit` = sum(Profit, na.rm = TRUE) / 1000000
  ) %>% 
  ggplot(aes(x = MAJ_DX_CAT, y = `Total Profit`, fill = MAJ_DX_CAT)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set1", direction = -1) +
  geom_hline(yintercept = 0, lty = 2, size = 1, color = "gray") +
  ylim(-2.15, 2.15) +
  labs(x = NULL, y = "$ Million", fill = NULL) +
  theme_bw() +
  theme(text = element_text(size = 16))

```


## 2

Total cost for top 10 disease entities in each group

```{r}
# description
top_10_description <- OMEGA %>%
  group_by(MAJ_DX_CAT, Description.1) %>%
  dplyr::summarise(
    Total_Patients = n(),
    `Total Cost (Avg.)` = mean(TOT_COST, na.rm = TRUE),
    `Total Pay (Avg.)` = mean(TOT_PAY, na.rm = TRUE)
    
  ) %>%
  group_by(MAJ_DX_CAT) %>%
  top_n(5, wt = Total_Patients) %>%
  arrange(-Total_Patients) %>% 
  ungroup()

top_10_description %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  flextable()
  
```


### Bar chart

```{r fig.height=4, fig.width=5}
top_10_description %>%
  pivot_longer(names_to = "key",
               values_to = "value",
               cols = c(4, 5)) %>%
  mutate(value = round(value, 2)) %>%
  ggplot(aes(
    x = reorder(Description.1, value),
    y = value / 1000,
    fill = key
  )) +
  geom_col(position = "dodge") +
  #ylim(0, 40) +
  scale_x_discrete(
    labels = function(x)
      str_wrap(x, width = 10)
  ) +
  coord_flip() +
  #geom_text(aes(label = value), position = position_dodge(width = 0.9), hjust = -0.1) +
  labs(x = "Procedure",
       y = "Average Cost ($1000)",
       fill = NULL,
       title = "Top 5 High Cost Procedures") +
  theme_minimal() +
  facet_wrap( ~ MAJ_DX_CAT, scales = "free_y") +
  theme(text = element_text(size = 14),
        legend.position = "top",
        plot.title = element_text(hjust = 0.5))
```

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>%
  summarise(
    `Total Cost` = sum(TOT_COST, na.rm = TRUE) / 1000000,
    `Total Pay` = sum(TOT_PAY, na.rm = TRUE) / 1000000
  )
```


## Total charge

```{r}
OMEGA %>% 
  ggplot(aes(x = TOT_CHARGE/1000)) +
  geom_histogram(fill = "darkblue") +
  labs(x = "Total Charges ($1000)", y = "Count") +
  theme_bw() +
  facet_wrap(~MAJ_DX_CAT)
#distribution = right skewed, most of the observations are skewed towards lower values for both ortho and cardio, only a handful of cases are greater than 50K

bb <- boxplot(TOT_CHARGE ~ MAJ_DX_CAT, data = OMEGA)

summary(bb$out)

OMEGA %>% 
  filter(!TOT_CHARGE %in% bb$out) %>% 
  ggplot(aes(x = TOT_CHARGE/1000)) +
  geom_histogram(fill = "darkblue") +
  labs(x = "Total Charges ($1000)", y = "Count") +
  theme_bw() +
  facet_wrap(~MAJ_DX_CAT)
# with outliers removed

library(rstatix)
OMEGA %>% 
  filter(!TOT_CHARGE %in% bb$out) %>% 
  select(MAJ_DX_CAT, TOT_CHARGE) %>% 
  group_by(MAJ_DX_CAT) %>% 
  get_summary_stats(type = "common")
```
Payer mix
```{r dpi = 400}
OMEGA %>%
  select(MAJ_DX_CAT, PRIM_PAY, TOT_CHARGE) %>%
  dplyr::count(MAJ_DX_CAT, PRIM_PAY) %>%
  group_by(MAJ_DX_CAT) %>%
  mutate(Fraction = round(n / sum(n), 2)) %>%
  ungroup() %>%
  ggplot(aes(
    x = reorder(PRIM_PAY, Fraction),
    y = Fraction,
    fill = MAJ_DX_CAT
  )) +
  geom_col(position = position_dodge()) +
  ylim(0, 0.6) +
  geom_text(aes(label = Fraction),
            position = position_dodge(width = 1),
            hjust = -0.1) +
  coord_flip() +
  labs(x = "Type of Payer",
       y = "% Reimbursement",
       fill = "Major Diagnosis Category",
       title = "Payer Mix - Cardiology vs. Orthopedics") +
  theme_minimal() 
#Medicare reimbursement much better for Cardio 
```


Profitability of patients from Cardiac and Orthopedic services
LOS outlier rates
Payer mix
Rates of emergent vs. elective care
Probable future demand given Omega's patient demographics for Cardiac and Orthopedic services (this is less of a coding/query question, and more of discussing what you can an infer from demographics and trends)

## Length of stay

```{r dpi = 400}
# boxplot
OMEGA %>% 
  select(ADM_DATE, DIS_DATE, LOS, MAJ_DX_CAT) %>% 
  group_by(MAJ_DX_CAT) %>% 
  mutate(Threshold = round(quantile(LOS, probs = c(0.99)), 0)) %>% 
  ggplot(aes(x = LOS, fill = MAJ_DX_CAT)) +
  geom_histogram() +
  geom_vline(aes(xintercept = Threshold, color = MAJ_DX_CAT), lty = 2, size = 0.5) +
  geom_text(aes(x = Threshold+3, y = 750, label = Threshold, color = MAJ_DX_CAT)) +
  labs(x = "Length of stay (Days)", y = "No. of Patients", fill = "Major Diagnosis Category", color = "99th Percentile", title = "Length of Stay Outliers - Cardiology vs. Orthopedics") +
  theme_minimal()
```





```{r}
# boxplot
OMEGA %>% 
  select(ADM_DATE, DIS_DATE, LOS, MAJ_DX_CAT) %>% 
  ggplot(aes(x = MAJ_DX_CAT, y = LOS)) +
  geom_boxplot(outlier.color = "red") +
  #geom_jitter(position = position_jitter(0.2), alpha = 0.5) +
  theme_bw()

# summary
OMEGA %>% 
  select(LOS, MAJ_DX_CAT) %>% 
  group_by(MAJ_DX_CAT) %>% 
  get_summary_stats() %>% 
  flextable()

# by top 10 disease entities
top_10_LOS <- OMEGA %>%
  select(MAJ_DX_CAT, Description.1, LOS) %>% 
  group_by(MAJ_DX_CAT, Description.1) %>%
  dplyr::summarise(
    Total_Patients = n(),
    LOS_mean = mean(LOS, na.rm = TRUE),
    LOS_sd = sd(LOS, na.rm = TRUE)
  ) %>%
  group_by(MAJ_DX_CAT) %>%
  top_n(5, wt = Total_Patients) %>%
  arrange(-Total_Patients) %>% 
  mutate(across(where(is.numeric), round, 2)) 

top_10_LOS %>% 
  flextable()
#Ask Rahul about 'outlier rate' and how to calculate
```

```{r fig.height=5, fig.width=8}
top_10_LOS %>%
  ggplot(aes(x = reorder(Description.1, LOS_mean),  y = LOS_mean)) +
  geom_col(fill = "darkblue") +
  coord_flip() +
  labs(x = NULL, y = "Average Length of Stay") +
  theme_bw() +
  facet_wrap( ~ MAJ_DX_CAT, scales = "free_y") +
  theme(text = element_text(size = 12))
```





# All Categories


## Data cleaning

### Unique patients


```{r}
omega_other_services <- omega_other_services %>% 
  distinct(PAT_NO, .keep_all = TRUE) %>% 
  filter(MAJ_DX_CAT != "**")


OMEGA <- OMEGA %>% 
  distinct(PAT_NO, .keep_all = TRUE) %>% 
  filter(MAJ_DX_CAT != "**")
```

## Profit in each group

### Profit


# bind all categories

```{r}
all_omega <- omega_other_services %>%
  select(PAT_NO, MAJ_DX_CAT, TOT_PAY, TOT_COST, LOS) %>%
  mutate(MAJ_DX_CAT = as.numeric(MAJ_DX_CAT)) %>%
  dplyr::bind_rows(OMEGA %>%
              select(PAT_NO, MAJ_DX_CAT, TOT_PAY, TOT_COST, LOS)) %>%
  mutate(Profit = TOT_PAY - TOT_COST) %>% 
  left_join(MDC_List %>% 
              mutate(MAJ_DX_CAT = as.numeric(MAJ_DX_CAT)), 
            by = "MAJ_DX_CAT")
```

all_omega %>% 
  count(MAJ_DX_CAT, Section)



### Boxplot

```{r}
all_omega %>% 
  mutate(Section_new = ifelse(!Section %in% c("Cardiology", "Orthopedic"), "Other", Section))  %>% 
  ggplot(aes(x = factor(MAJ_DX_CAT), y = Profit, color = Section_new)) +
  geom_boxplot() +
  labs(x = "Major Categories", color = NULL, title = "Profit Distribution in all Major Diagnosis Categories") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
```


### Fraction of profit

```{r}
AA <- all_omega %>%
  group_by(MAJ_DX_CAT, Section) %>%
  summarise(Profit = mean(Profit, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(Fraction = round((Profit/sum(Profit))*100), 1) %>% 
  mutate(Section_new = ifelse(!Section %in% c("Cardiology", "Orthopedic"), "Other", Section)) 

AA %>% 
  ggplot(aes(x = reorder(Section, Fraction), y = Fraction, fill = Section_new)) +
  geom_col(position = "dodge") +
  coord_flip() +
  geom_text(aes(label = Fraction), hjust = 1, size = 3) +
  labs(x = "Major Diagnostic Category", y = "Profit (%)", fill = NULL, title = " Percentage of Profit by Major Diagnostic Category ") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
```


## Length of Stay

### Above 99th percentile


```{r}
LOS_OR <- all_omega %>% 
  select(-c(TOT_PAY, TOT_COST, Profit)) %>% 
  group_by(MAJ_DX_CAT, Section) %>% 
  mutate(Threshold = quantile(LOS, probs = c(0.99), na.rm = TRUE)) %>% 
  mutate(Outlier = ifelse(LOS > Threshold, "Yes", "No")) %>% 
  group_by(MAJ_DX_CAT, Section, Outlier) %>% 
  summarise(N = n()) %>% 
  group_by(MAJ_DX_CAT, Section) %>% 
  mutate(Fraction = round(N/sum(N)*100, 1)) %>% 
  filter(Outlier == "Yes") %>% 
  ungroup() %>% 
  arrange(-Fraction) %>% 
  mutate(Section_new = ifelse(!Section %in% c("Cardiology", "Orthopedic"), "Other", Section))

pp <- LOS_OR %>% 
  ggplot(aes(x = reorder(Section, Fraction), y = Fraction, fill = Section_new)) +
  geom_col(position = "dodge") +
  coord_flip() +
  geom_text(aes(label = Fraction), hjust = -0.25, size = 3) +
  labs(x = "Major Diagnostic Category", y = "Outlier (%)", fill = NULL, title = "Outlier Rates in All Major Diagnostic Categories") +
  theme_bw() +
  theme(text = element_text(size = 13),
        legend.position = "none")

pp

ggsave("~/Downloads/Outlier_Rate.png", pp, dpi = 400)
  
```
```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>% 
  dplyr::summarise(N = n()) %>% 
  flextable()
```



# Questions

## 1

Top disease entities treated by each group (Cardiac and Orthopedic) as well as the number of patients in each of these disease entities

### Number of patients in each group





```{r}
OMEGA %>% 
  group_by(MAJ_DX_CAT, Description.1) %>% 
  dplyr::summarise(N = n()) %>% 
  group_by(MAJ_DX_CAT) %>% 
  top_n(5) 
```



## Total charges in each group?

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>%
  summarise(
    `Total Cost` = sum(TOT_COST, na.rm = TRUE) / 1000000,
    `Total Pay` = sum(TOT_PAY, na.rm = TRUE) / 1000000
  ) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  flextable()
```


## Bar chart

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>%
  dplyr::summarise(
    `Total Cost` = sum(TOT_COST, na.rm = TRUE) / 1000000,
    `Total Revenue` = sum(TOT_PAY, na.rm = TRUE) / 1000000
  ) %>% 
  pivot_longer(!MAJ_DX_CAT) %>% 
  ggplot(aes(x = MAJ_DX_CAT, y = value, fill = name)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "$ Million", fill = NULL) +
  theme_bw() +
  theme(text = element_text(size = 16))
  geom_text(aes(label = n), hjust = 1.3, size = 3.5)
#slight loss in ortho
#profitability 

```

## Profitibility

### Histograms

```{r}
OMEGA %>%
  mutate(Profit = TOT_PAY - TOT_COST) %>% 
  ggplot(aes(x = Profit)) +
  geom_histogram(fill = "orange") +
  geom_vline(xintercept = 0, lty = 2, size = 1, color = "red") +
  theme_test() +
  facet_wrap(~MAJ_DX_CAT, scales = "free_x")
  geom_text(aes(label = n), hjust = 1.3, size = 3.5)
```


```{r}
p <- OMEGA %>%
  mutate(Profit = TOT_PAY - TOT_COST) %>% 
  group_by(MAJ_DX_CAT) %>%
  dplyr::summarise(
    `Total Profit` = sum(Profit, na.rm = TRUE) / 1000000
  ) %>% 
  ggplot(aes(x = MAJ_DX_CAT, y = `Total Profit`, fill = MAJ_DX_CAT)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set1", direction = -1) +
  geom_hline(yintercept = 0, lty = 2, size = 1, color = "gray") +
  ylim(-2.15, 2.15) +
  labs(x = NULL, y = "$ Million", fill = NULL) +
  theme_bw() +
  theme(text = element_text(size = 16))
 
```


## 2

Total cost for top 10 disease entities in each group


```{r}
# description
top_10_description <- OMEGA %>%
  group_by(MAJ_DX_CAT, Description.1) %>%
  dplyr::summarise(
    Total_Patients = n(),
    `Total Cost (Avg.)` = mean(TOT_COST, na.rm = TRUE),
    `Total Pay (Avg.)` = mean(TOT_PAY, na.rm = TRUE)
    
  ) %>%
  group_by(MAJ_DX_CAT) %>%
  top_n(5, wt = Total_Patients) %>%
  arrange(-Total_Patients) %>% 
  ungroup()

top_10_description %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  flextable()

```

### Bar chart

```{r fig.height=4, fig.width=5}
top_10_description %>%
  pivot_longer(names_to = "key", values_to = "value", cols = c(4, 5)) %>% 
  ggplot(aes(x = reorder(Description.1, value),  y = value / 1000, fill = key)) +
  geom_col(position = "dodge") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  coord_flip() +
  labs(x = NULL, y = "Average Cost ($1000)", fill = NULL) +
  theme_bw() +
  facet_wrap( ~ MAJ_DX_CAT, scales = "free_y") +
  theme(text = element_text(size = 14), 
        legend.position = "top")
  
  
```

```{r}
OMEGA %>%
  group_by(MAJ_DX_CAT) %>%
  summarise(
    `Total Cost` = sum(TOT_COST, na.rm = TRUE) / 1000000,
    `Total Pay` = sum(TOT_PAY, na.rm = TRUE) / 1000000
  )
```


## Total charge

```{r}
OMEGA %>% 
  ggplot(aes(x = TOT_CHARGE/1000)) +
  geom_histogram(fill = "darkblue") +
  labs(x = "Total Charges ($1000)", y = "Count") +
  theme_bw() +
  facet_wrap(~MAJ_DX_CAT)
#distribution = right skewed, most of the observations are skewed towards lower values for both ortho and cardio, only a handful of cases are greater than 50K

bb <- boxplot(TOT_CHARGE ~ MAJ_DX_CAT, data = OMEGA)

summary(bb$out)

OMEGA %>% 
  filter(!TOT_CHARGE %in% bb$out) %>% 
  ggplot(aes(x = TOT_CHARGE/1000)) +
  geom_histogram(fill = "darkblue") +
  labs(x = "Total Charges ($1000)", y = "Count") +
  theme_bw() +
  facet_wrap(~MAJ_DX_CAT)
# with outliers removed

library(rstatix)
OMEGA %>% 
  filter(!TOT_CHARGE %in% bb$out) %>% 
  select(MAJ_DX_CAT, TOT_CHARGE) %>% 
  group_by(MAJ_DX_CAT) %>% 
  get_summary_stats(type = "common")
```
Payer mix
```{r}
OMEGA %>% 
  select(MAJ_DX_CAT, PRIM_PAY, TOT_CHARGE) %>% 
  dplyr::count(MAJ_DX_CAT, PRIM_PAY) %>% 
  ggplot(aes(x = reorder(PRIM_PAY, n),  y = n, fill = MAJ_DX_CAT)) +
  geom_col(position = position_dodge()) +
  coord_flip() +
  labs(x = "Type of Payer", y = "Number of Patients", fill = NULL, title = "") +
  theme_bw() 
#Medicare reimbursement much better for Cardio 
```


Profitability of patients from Cardiac and Orthopedic services
LOS outlier rates

Payer mix
Rates of emergent vs. elective care
Probable future demand given Omega's patient demographics for Cardiac and Orthopedic services (this is less of a coding/query question, and more of discussing what you can an infer from demographics and trends)

## Length of stay

```{r}
# boxplot
OMEGA %>% 
  select(ADM_DATE, DIS_DATE, LOS, MAJ_DX_CAT) %>% 
  ggplot(aes(x = MAJ_DX_CAT, y = LOS)) +
  geom_boxplot(outlier.color = "red") +
  #geom_jitter(position = position_jitter(0.2), alpha = 0.5) +
  theme_bw()

# summary
OMEGA %>% 
  select(LOS, MAJ_DX_CAT) %>% 
  group_by(MAJ_DX_CAT) %>% 
  get_summary_stats() %>% 
  flextable()

# by top 10 disease entities
top_10_LOS <- OMEGA %>%
  select(MAJ_DX_CAT, Description.1, LOS) %>% 
  group_by(MAJ_DX_CAT, Description.1) %>%
  dplyr::summarise(
    Total_Patients = n(),
    LOS_mean = mean(LOS, na.rm = TRUE),
    LOS_sd = sd(LOS, na.rm = TRUE)
  ) %>%
  group_by(MAJ_DX_CAT) %>%
  top_n(5, wt = Total_Patients) %>%
  arrange(-Total_Patients) %>% 
  mutate(across(where(is.numeric), round, 2)) 

top_10_LOS %>% 
  flextable()
#Ask Rahul about 'outlier rate' and how to calculate
```

- Which section has more LOS outliers?



```{r fig.height=5, fig.width=8}
top_10_LOS %>%
  ggplot(aes(x = reorder(Description.1, LOS_mean),  y = LOS_mean)) +
  geom_col(fill = "darkblue") +
  coord_flip() +
  labs(x = NULL, y = "Average Length of Stay") +
  theme_bw() +
  facet_wrap( ~ MAJ_DX_CAT, scales = "free_y") +
  theme(text = element_text(size = 12))
```







